Generating Datasets about Building Designs \\Based on a Multilayer Building Ontology

### 2015-04-29 article-02-main.tex
Reduced number of headings

\section{Generating ifcOWL Ontologies From IFC Specifications}
\label{sec:ifcOWL}
This section describes the derivation of the ifcOWL ontologies from the IFC specification. Since there are different options for the conversion, the choices are presented as conversion principles with justiciations.

@@

\begin{principle}[Simple OWL profiles]
\ifcowl{} should be compatible with as simple OWL 2 profiles as possible (to support a maximum variety of reasoners and other tools). 
\end{principle}

The first layer \ifcsimple{} can be specified to be compatible with any OWL 2 DL profile: OWL 2 EL, OWL 2 QL or OWL 2 RL. The other layers \ifcstandard{} and \ifcextended{} are compatible with OWL 2 RL\footnote{Our WebIFC converter tool allows forced choice of other OWL profiles for any layer: OWL Lite, OWL DL, OWL Full, and OWL 2 Full}. OWL 2 RL needed because it -- unlike OWL 2 EL or QL -- supports 
union data\-types (see \ref{subsec:ifcOWL-select-types}), enumerations of individuals (see \ref{subsec:ifcOWL-enum-types}), and anonymous individuals (see \ref{subsec:ifcRDF-naming-individuals}) required by the IFC schema. \cite{w3c:owl2-profiles}

\begin{principle}[No domain and range restrictions]
Property names should be kept simple and therefore domain and range constraints (specified by \name{rdfs:domain} and \name{rdfs:range}) should not be included in \ifcowl{} ontologies.
\end{principle}

@@

\begin{principle}[Namespaces]
All constructs related to EXPRESS, STEP or IFC specifications must be belong to namespaces shown below with prefixes \name{expr:}, \name{step:} and \name{ifc:} accordingly (instead of \name{IFCXXX} there should be the name of IFC version, e.g. \name{IFC4_ADD1}):
%\name{IFC2x3} or \name{IFC4}):
%At this moment the namespaces are defined as below (instead of \name{IFCXXX} there should be IFC versions, e.g. 
%\name{IFC2x3} or \name{IFC4}):

@@

@@

\subsection{Components of an EXPRESS schema}
The IFC specification is written as an EXPRESS schema. An EXPRESS schema defines series of data\-types, functions and rules by using EXPRESS data definition language specified in ISO10303-11 \cite{wiki:express,noauthor:ifc-guide}.

There are six groups of data\-types: simple data\-types, entity data\-types, enumeration data\-types, defined (declared) data\-types, select data\-types and aggregation data\-types. All types must be declared obviously by using constructs TYPE and ENTITY except the built-in simple data\-types. Definitions and characteristics of data\-types are considered below. EXPRESS schemata also contain definitions of functions, rules and type constraints most of which cannot be reflected in OWL 2 at all.

@@

\subsection{Simple Datatype Declarations}
\label{subsec:ifcOWL-simple-type}

Simple data\-types built in EXPRESS are: \name{STRING}, \name{BI\-NA\-RY}, \name{IN\-TE\-GER}, \name{REAL}, \name{NUM\-BER}, \name{BOO\-LEAN} and \name{LO\-GI\-CAL}. However, type \name{LOGICAL} which has three possible values \name{TRUE}, \name{FALSE} and \name{UNKNOWN} is considered by us as an enumeration data\-type (see \ref{subsec:ifcOWL-enum-types}). The same to type \name{BOOLEAN} which is like a primitive boolean data\-type but semantically close and met with the same frequency in IFC4 as \name{LOGICAL}.

\begin{principle}[Simple data\-types]
All simple data\-types except \emph{BOOLEAN} and \emph{LOGICAL} are declared as the same as their most similar primitive XSD data\-types which are supported by OWL 2 \cite{w3c:owl2-profiles} and are among preferred standard data\-types in Linked Data Platform \cite{w3c:ldp-best-practices}:

@@

`Most similar XSD data\-types' mean the types from XSD schema that conform to the value spaces of the original EXPRESS and IFC data\-types better than other ones. For instance, type \name{xsd:double} is more similar to \name{REAL} than \name{xsd:decimal} as it is an IEEE 64-bit floating-point data\-type and supports scientific notation \cite{w3c:xsd,datapic:xsd}. It is recommended to use XSD data\-type names directly when they are needed.

\begin{principle}[Simple data\-types with sizes, ifcOWL layers: all]If \name{STRING} and \name{BINARY} data\-types are specified with a size parameter, i.e. \name{STRING (max\-Length\-In\-Bytes)} or \name{BINARY(size\-In\-Bits)}, then in layers \simple{} and \standard{} they are defined as equivalent to data\-types without size. But in layer \extended{}, the size parameter must be declared by using property \name{owl:onDatatype} and data\-type facets \name{xsd:maxLength} or \name{xsd:length} \cite{w3c:owl2-syntax,w3c:xmlschema11-2}, for instance:

% \begin{lstlisting}[caption={Size constraint of simple datatypes},label=lst:ifcOWL-simple-types-with-size]
\begin{lstlisting}
# only for Simple, Standard layers
expr:STRING22 owl:sameAs expr:STRING .

# only for Extended layer
expr:STRING22 owl:equivalentClass
    [   a rdfs:Datatype ;
        owl:onDatatype  xsd:string ;
        owl:withRestrictions ( [ xsd:length "22"^^xsd:integer ] ) ] .
\end{lstlisting}
\end{principle}

@@

\subsection{Enumeration Datatype Declarations}
\label{subsec:ifcOWL-enum-types}

An enumeration data\-type represents a set of possible fixed values.

\begin{principle}[Enumeration data\-types]
Each enumeration data\-type in \textbf{ifc\-OWL} is defined as an \name{owl:Class} which is a subclass of \name{expr:\-Enum\-Class}. All possible values of enumeration data\-types are defined as named individuals. Enumeration data\-type memberships are defined by means of \name{rdf:type} in the \simple{} layer, and by means of \name{owl:oneOf} in \standard{} and \extended{} layers as shown below:
% \begin{lstlisting}[caption={Enumeration data\-types}, label=lst:ifcOWL-enum-types]
\begin{lstlisting}
# only for Simple layer
ifc:IfcAssemblyPlaceEnum a owl:Class ; rdfs:subclassOf expr:EnumClass .
ifc:SITE, ifc:FACTORY, ifc:NOTDEFINED rdf:type ifc:IfcAddressTypeEnum .

# only for Standard, Extended layers
ifc:IfcAssemblyPlaceEnum a owl:Class ;  rdfs:subclassOf expr:EnumClass ;
    owl:oneOf ( ifc:SITE ifc:FACTORY ifc:NOTDEFINED ) .
\end{lstlisting}
\end{principle}

The reason of two different approaches is that enumeration with more than one individual or literal are not supported in OWL 2 EL or QL. In both cases enumeration values are defined as individuals of one class so in terms of reasnoning logic they are equivalent. The enumeration values are defined as individuals because OWL 2 RL supports \name{Object\-One\-Of} to define enumeration of individuals, but not \name{Data\-One\-Of} \cite{w3c:owl2-profiles}. The fact that some values such as \name{ifc:NOT\-DE\-FI\-NED} belong to different enumeration data\-types should not cause any problems since all enumeration values in ifc\-RDF datasets are defined together with a concrete enumeration data\-type (see \ref{subsec:ifcRDF-enum-values}). It is also very easy to get all possible values of an enumeration data\-type by using simple SPARQL queries.

As mentioned above, simple data\-types \name{BOOLEAN} and \name{LOGICAL} are considered by us as enumeration data\-types. But these types and their values must be defined in the EXPRESS namespace, e.g. \name{expr:LOGICAL}, \name{expr:TRUE}, \name{expr:FALSE} or \name{expr:UNKNOWN}.

@@

\subsection{Select Datatype Declarations}
\label{subsec:ifcOWL-select-types}

A select data\-type represents an union of other data\-types. Union members of a select data\-type usually are either entity data\-types, or defined data\-types all together. But there are a few exceptions, for instance, type \name{Ifc\-Trim\-ming\-Se\-lect} is an union of entity data\-type \name{Ifc\-Car\-te\-sian\-Point} and defined data\-type \name{Ifc\-Pa\-ra\-me\-ter\-Va\-lue}.

\begin{principle}[Select data\-types]
A select data\-type must be converted to an \name{owl:Class} which is a subclass of \name{expr:\-Select\-Class}. Union-member data\-types are defined by means of \name{rdfs:\-subclassOf} in the \simple{} layer, and by means of \name{owl:unionOf} in \standard{} and \extended{} layers as shown in the listing below.
\end{principle}

%\begin{lstlisting}[caption={Select data\-types}, label=lst:ifcOWL-select-types]
\begin{lstlisting}
# only for Simple layer
ifc:IfcTrimmingSelect a owl:Class ; rdfs:subclassOf expr:SelectClass .
ifc:IfcCartesianPoint, ifc:IfcParameterValue
    rdfs:subclassOf ifc:IfcTrimmingSelect .

# only for Standard, Extended layers
ifc:IfcTrimmingSelect a owl:Class ; rdfs:subclassOf expr:SelectClass ;
    owl:unionOf ( ifc:IfcCartesianPoint ifc:IfcParameterValue ) ] .
\end{lstlisting}

Similarly to the above case, two different approaches are used because property \name{owl:unionOf} supported only in OWL 2 RL, but not in OWL 2 EL and OWL 2 QL \cite{w3c:owl2-profiles}.

@@

\subsection{Defined Datatype Declarations}
\label{subsec:ifcOWL-defined-types}

A defined (declared, named) data\-type is usually specified as equivalent to a simple data\-type or another defined data\-type, with or without a additional constraint. For instance, type \name{Ifc\-Length\-Mea\-sure} is defined as equal to \name{REAL}, and type \name{Ifc\-Po\-si\-ti\-ve\-Length\-Mea\-su\-re} is a set of positive \name{Ifc\-Length\-Mea\-sure} values.

\begin{principle}[Defined data\-types]
If the defined data\-type is based on a simple data\-type then in ifcOWL it is declared as a \name{owl:Class} which is a subclass of \name{expr:\-Defined\-Class} has a single property \name{rdf:value} with that simple data\-type. If the defined data\-type is based on another defined data\-type then it is declared as a subclass of the \name{owl:Class} of that defined data\-type, for instance:

%\begin{lstlisting}[caption={Defined data\-types}, label=lst:ifcOWL-defined-types]
\begin{lstlisting}[aboveskip=3pt]
ifc:IfcLengthMeasure a owl:Class ; rdfs:subClassOf expr:DefinedClass ;
    rdfs:subClassOf [ a owl:Restriction; owl:onProperty rdf:value;
                      owl:allValuesFrom xsd:double ] .

ifc:IfcPositiveLengthMeasure a owl:Class ;
    rdfs:subClassOf ifc:IfcLengthMeasure .
\end{lstlisting}
\end{principle}

Defined data\-types are not declared as a \name{rdfs:Datatype}, but \name{owl:Class}es which wrap \name{rdfs:Datatype} values because in contract to simple data types they can be used for declaration of select data\-types (see \ref{subsec:ifcOWL-select-types}).

Defined data\-type \name{Ifc\-Time\-Stamp} is declared as equivalent to \name{INTEGER} but represents so-called Unix time value that can be easily converted to \name{xsd:date\-Time}. So it wraps a \name{xsd:date\-Time} value instead of \name{xsd:integer}.

@@

\subsection{Entity Datatype Declarations}
\label{subsec:ifcOWL-entity-types}

Entity data\-types are identical to classes in object-oriented programming languages. Each type may have not more than one super\-type that it inherits attributes from. An attribute is defined with a data\-type, a minimum and a maximum cardinalities, can be marked as optional.

\begin{principle}[Entity data\-types]An entity data\-type is converted to an \name{owl:\-Class} which is subclass of \name{expr:\-Entity\-Class} or its superclass. Entity attributes are defined by means of property \name{owl:on\-Pro\-per\-ty}. If an attribute is optional then its minimum qualified cardinality is 0, otherwise -- 1. Maximum qualified of all attributes are always 1. In case if an attribute is a list or a set of some data\-type then its type is considered as an external aggregated data\-type, for example:
%\begin{lstlisting}[caption={Entity data\-types}, label=lst:ifcOWL-defined-types]
\begin{lstlisting}
ifc:IfcProperty a owl:Class ; rdfs:subClassOf expr:EntityClass .

ifc:IfcComplexProperty a owl:Class ; rdfs:subClassOf ifc:IfcProperty ;
    rdfs:subClassOf [ a owl:Restriction; owl:onProperty ifc:coordinates;
                      owl:allValuesFrom ifc:LIST_1_3_IfcLengthMeasure ] ;
    # only for Extended layer
    rdfs:subClassOf [ a owl:Restriction; owl:onProperty ifc:coordinates;
                      owl:cardinality "1"^^xsd:integer ] .

@@

\subsection{Aggregated Type Declarations}
\label{subsec:ifcOWL-aggregated-types}

Aggregation data\-types may be unordered (\name{SET} and \name{BAG}) or  ordered (\name{LIST} and \name{ARRAY}). \name{BAG} may contain a particular value more than once, unlike \name{SET}. \name{ARRAY} is specified with starting and ending indices while other types are specified with min and max cardinalities.

In order to declare these types, new classes were created in a similar way as Ordered List Ontoloty \cite{olo:ontology}:
\begin{itemize}
\item Class \name{expr:\-Collec\-tion\-Class} has four functional properties: \name{expr:isOrdered}, \name{expr:item\-Type}, \name{expr:\-size}, \name{expr:\-start\-Index}, \name{expr:\-end\-Index} and a nonfunctional property \name{expr:slot} with type \name{expr:Slot}.
\item Class \name{expr:Slot} has two functional properties \name{expr:item} (the actual collection item value), \name{expr:index} and also optional functional properties \name{expr:previous}, \name{expr:next}.
\item Classes \name{expr:SetClass}, \name{expr:ListClass}, \name{expr:ArrayClass} and \name{expr:BagClass} are subclasses of \name{expr:\-Collec\-tion\-Class}.
\end{itemize}

Below is a fragment of definition of type \name{SET [2:?] OF IfcProfileDef}:

\begin{lstlisting}
# all layers:
 ifc:SET_2_UNDEFINED_OF_IfcProfileDef a owl:Class ;
    rdfs:subClassOf ifc:SET_2_UNDEFINED , ifc:SET_OF_IfcProfileDef .

 ifc:SET_2_UNDEFINED, SET_OF_IfcProfileDef a owl:Class ;
    rdfs:subClassOf expr:SetClass .

# only for layer Extended:
 ifc:SET_2_UNDEFINED rdfs:subClassOf rdfs:subClassOf [ a owl:Restriction ;
    owl:onProperty expr:slot ; owl:minCardinality "2"^^xsd:integer ] .

 ifc:SET_OF_IfcProfileDef rdfs:subClassOf [ a owl:Restriction ;
    owl:onProperty expr:slot ; owl:allValuesFrom ifc:SLOT_OF_IfcProfileDef ].

 ifc:SLOT_OF_IfcProfileDef rdfs:subClassOf [ a owl:Restriction ; owl:onProperty expr:item ; owl:allValuesFrom ifc:IfcProfileDef ] .
\end{lstlisting}

@@

\section{Generating ifcRDF Datasets From IFC Data}
\label{sec:ifcRDF}

\subsection{Dataset Metadata}
\label{subsec:ifcRDF-metadata}

@@

\begin{lstlisting}
:GUID_1916794F-5605-4499-9AB0-F155DE8D3B6C
    a               ifc:IfcPropertySet ; # entity type
    ifc:globalId    [ a ifc:IfcGloballyUniqueId ; # defined type
                     rdf:value "0P5dbFLWL4cPgmyLNUZJji"^^xsd:string ] ;
    ifc:hasProperties  _:b11 .

_:b11 a ifc:SET_1_UNLIMITED_OF_IfcProperty ; # entity type
    expr:size "3"^^xsd:integer ;
    expr:itemType ifc:IfcProperty ;
    expr:slot _:b12, _:b13, _:b14 .

_:b12 a ifc:SLOT_OF_IfcProperty ;
    expr:item [ a ifc:IfcPropertySingleValue ; # entity type
                ifc:name "initial_GUID"^^xsd:string ;
                ifc:nominalValue  _:b15 ] .
\end{lstlisting}

>>>>

\section{Principles of Generating ifcOWL Ontologies And ifcRDF Datasets}
\label{sec:ifcOWL}
This section describes the derivation of the ifcOWL ontologies from the IFC specification. Since there are different options for the conversion, the choices are presented as conversion principles with justiciations.

@@

\begin{principle}%[Simple OWL profiles]
\ifcsimple{} should be compatible with as simple OWL 2 profiles as possible (to support a maximum variety of reasoners and other tools). 
\end{principle}

The first layer \ifcsimple{} can be specified to be compatible with any OWL 2 DL profile: OWL 2 EL, OWL 2 QL or OWL 2 RL. The other layers \ifcstandard{} and \ifcextended{} are compatible with OWL 2 RL. OWL 2 RL needed because it -- unlike OWL 2 EL or QL -- supports such necessary things like enumerations of individuals, union classes, disjoint classes, and also anonymous individuals. \cite{w3c:owl2-profiles}

\begin{principle}%[No domain and range restrictions]
Property names should be kept simple and therefore domain and range constraints (specified by \name{rdfs:domain} and \name{rdfs:range}) should not be included in \ifcowl{} ontologies.
\end{principle}

@@

\begin{principle}%[Namespaces]
All constructs related to EXPRESS, STEP or IFC specifications must be belong to namespaces shown below with prefixes \name{expr:}, \name{step:} and \name{ifc:} accordingly (instead of \name{IFCXXX} there should be the name of IFC version, e.g. \name{IFC4\_ADD1}):
%\name{IFC2x3} or \name{IFC4}):
%At this moment the namespaces are defined as below (instead of \name{IFCXXX} there should be IFC versions, e.g. 
%\name{IFC2x3} or \name{IFC4}):

@@

\begin{principle}All definitions of functions, rules, type constraints and derived attributes specified in EXPRESS after keywords \name{FUNCTION}, \name{RULE}, \name{WHERE} and \name{RULE} accordingly should be ignored.
\end{principle}

\TODO{Explain why functions, rules, etc. should be ignored}

@@

\subsection{Datatype Declarations}
\label{subsec:ifcOWL-types}

\TODO{\textit{Write about EXPRESS in the Introduction section}:\\
The IFC specification is written as an EXPRESS schema. An EXPRESS schema defines series of data\-types, functions and rules by using EXPRESS data definition language specified in ISO10303-11 \cite{wiki:express,noauthor:ifc-guide}.}

An EXPRESS schema includes six groups of data\-types: built-in simple data\-types, entity data\-types, enumeration data\-types, defined data\-types, select data\-types and aggregation data\-types.

@@

\subsubsection{Simple Datatypes} built in EXPRESS are: \name{STRING}, \name{BI\-NA\-RY}, \name{IN\-TE\-GER}, \name{REAL}, \name{NUM\-BER}, \name{BOO\-LEAN} and \name{LO\-GI\-CAL}. Types \name{STRING} and \name{BINARY} data\-types may be specified with a size parameter: \name{STRING (max\-Length\-In\-Bytes)} or \name{BINARY(size\-In\-Bits)}.

\begin{principle}%[Simple data\-types]
A simple data\-type (except \emph{BOOLEAN} and \emph{LOGICAL}) should be declared in \ifcowl{} as \name{owl:sameAs} of its most similar primitive XSD data\-type among all data\-types supported by any OWL 2 profile \cite{w3c:owl2-profiles} and preferred in Linked Data Platform \cite{w3c:ldp-best-practices}:

@@

Type \name{LOGICAL} -- with its three possible values \name{TRUE}, \name{FALSE} and \name{UNKNOWN} -- is considered as an enumeration data\-type. The same to \name{BOOLEAN} because OWL 2 EL and QL do not support type \name{xsd:boolean} \cite{w3c:owl2-profiles}.

Statement `most similar XSD data\-type' means the type from XSD schema that conform to the value spaces of the EXPRESS simple data\-type better than any other type. For instance, type \name{xsd:double} is more similar to \name{REAL} than \name{xsd:decimal} as it is an IEEE 64-bit floating-point data\-type and supports scientific notation \cite{w3c:xsd,datapic:xsd}. When it is needed to refer to an EXPRRESS simple data\-type, we recommend to use its XSD equivalent type directly.

\begin{principle}%[Simple data\-types with sizes]
Size parameters of types \name{STRING} and \name{BINARY} (if any) should be ignored in layers \simple{} and \standard{} and should be declared obviously -- by means of OWL 2 data\-type facets -- in layer \extended{}, for instance:

% they are defined as equivalent to data\-types without size. But in layer \extended{}, the size parameter must be declared by using property \name{owl:onDatatype} and data\-type facets \name{xsd:maxLength} or \name{xsd:length} \cite{w3c:owl2-syntax,w3c:xmlschema11-2}, for instance:

% \begin{lstlisting}[caption={Size constraint of simple datatypes},label=lst:ifcOWL-simple-types-with-size]
\begin{lstlisting}
# only for Simple, Standard layers
  expr:STRING22 owl:sameAs xsd:string .
# only for Extended layer
  expr:STRING22 owl:equivalentClass [   a rdfs:Datatype ;
    owl:onDatatype  xsd:string; owl:withRestrictions ( [ xsd:length 22 ] ) ].
\end{lstlisting}
\end{principle}

@@

\subsubsection{Enumeration Datatypes} represent fixed sets of possible values.

\begin{principle}%[Enumeration data\-types]
Each enumeration data\-type should be defined in \ifcowl{} as an \name{owl:Class} which is a subclass of \name{expr:\-Enum\-Class}. Enumeration values must be defined as named individuals of the enumeration class but in different ways in different layers as shown below:

% \begin{lstlisting}[caption={Enumeration data\-types}, label=lst:ifcOWL-enum-types]
\begin{lstlisting}
# only for Simple layer
  ifc:IfcAssemblyPlaceEnum a owl:Class ; rdfs:subclassOf expr:EnumClass .
  ifc:SITE, ifc:FACTORY, ifc:NOTDEFINED rdf:type ifc:IfcAddressTypeEnum .
# only for Standard, Extended layers
  ifc:IfcAssemblyPlaceEnum a owl:Class ;  rdfs:subclassOf expr:EnumClass ;
    owl:oneOf ( ifc:SITE ifc:FACTORY ifc:NOTDEFINED ) .
\end{lstlisting}
\end{principle}

Two different approaches are used because enumerations with more than one individual or literal are not supported in OWL 2 EL or QL, like enumerations of literals in OWL 2 RL. The fact that some values -- for instance, \name{ifc:NOT\-DE\-FI\-NED} -- belong to different enumeration data\-types should not cause any problems since any enumeration value is always specified with an exact type.

As mentioned above, simple data\-types \name{BOOLEAN} and \name{LOGICAL} are considered as enumeration data\-types. But these types and their values must be defined in the EXPRESS-related namespace, e.g. \name{expr:LOGICAL}, \name{expr:TRUE}, \name{expr:UNKNOWN}.

@@

\subsubsection{Select Datatypes} represent unions of other data\-types. Particular members of a select data\-type can be entity, defined or select data\-types. For instance, type \name{Ifc\-Trim\-ming\-Se\-lect} is an union of entity data\-type \name{Ifc\-Car\-te\-sian\-Point} and defined data\-type \name{Ifc\-Pa\-ra\-me\-ter\-Va\-lue}.

\begin{principle}%[Select data\-types]
A select data\-type must be converted to an \name{owl:Class} which is a subclass of \name{expr:\-Select\-Class}. Its particular data\-types are defined as subclasses of the select class, but in different ways depended on the layer:
\end{principle}

%\begin{lstlisting}[caption={Select data\-types}, label=lst:ifcOWL-select-types]
\begin{lstlisting}
# only for Simple layer
  ifc:IfcTrimmingSelect a owl:Class ; rdfs:subclassOf expr:SelectClass .
  ifc:IfcCartesianPoint, ifc:IfcParameterValue
    rdfs:subclassOf ifc:IfcTrimmingSelect .
# only for Standard, Extended layers
  ifc:IfcTrimmingSelect a owl:Class ; rdfs:subclassOf expr:SelectClass ;
    owl:unionOf ( ifc:IfcCartesianPoint ifc:IfcParameterValue ) ] .
\end{lstlisting}

Two different approaches are used because property \name{owl:unionOf} is supported only in OWL 2 RL, but not in OWL 2 EL and QL \cite{w3c:owl2-profiles}.

@@

\subsubsection{Defined (Declared, Named) Datatypes} are usually specified as equivalent or subclass of simple data\-types or other defined data\-types. For instance, type \name{Ifc\-Length\-Mea\-sure} is equal to \name{REAL}, while type \name{Ifc\-Po\-si\-ti\-ve\-Length\-Mea\-su\-re} is a subset of \name{Ifc\-Length\-Mea\-sure} with some constraint.

\begin{principle}%[Defined data\-types]
A defined data\-type based on a simple data\-type should be declared as an \name{owl:Class} which is a subclass of \name{expr:\-Defined\-Class} and has a single property \name{rdf:value} with that simple data\-type. A defined data\-type based on another defined data\-type should be declared as a subclass of the base defined data\-type, for instance:

%\begin{lstlisting}[caption={Defined data\-types}, label=lst:ifcOWL-defined-types]
\begin{lstlisting}
ifc:IfcLengthMeasure a owl:Class ; rdfs:subClassOf expr:DefinedClass ;
    rdfs:subClassOf [ a owl:Restriction; owl:onProperty rdf:value;
                      owl:allValuesFrom xsd:double ] .
ifc:IfcPositiveLengthMeasure rdfs:subClassOf ifc:IfcLengthMeasure .
\end{lstlisting}
\end{principle}

Defined data\-types are defined as \name{owl:Class}es instead of \name{rdfs:Datatype}s because they can be merged with other \name{owl:Class}es for creating select data\-types. In addition, different defined data\-types wrapping the same \name{rdfs:Datatype} are not the same. For example, \name{IfcLengthMeasure} is different to \name{IfcMassMeasure} although they both are \name{REAL} and belong to select class \name{IfcMeasureValue}.

In principle type \name{Ifc\-Time\-Stamp} must wrap an integer value which represents so-called Unix time. But we decided this value should have type \name{xsd:date\-Time}, and all values of \name{Ifc\-Time\-Stamp} must be converted to a standard datetime values.

@@

\subsubsection{Entity Datatypes} are identical to classes in object-oriented programming languages. Each type may have one super\-type. An attribute is defined with a data\-type, cardinalities and can be marked as optional.

\begin{principle}%[Entity data\-types]
An entity data\-type should be converted to an \name{owl:\-Class} which is subclass of its superclass (if any) or \name{expr:\-Entity\-Class}. Entity attribute cardinalities can be only from 0 to 1, even if this attribute is a set of values. In this case the attribute type is an external aggregated data\-type, for example:

%\begin{lstlisting}[caption={Entity data\-types}, label=lst:ifcOWL-defined-types]
\begin{lstlisting}
ifc:IfcProperty a owl:Class ; rdfs:subClassOf expr:EntityClass .
ifc:IfcComplexProperty a owl:Class ; rdfs:subClassOf ifc:IfcProperty ;
    rdfs:subClassOf [ a owl:Restriction; owl:onProperty ifc:coordinates;
                      owl:allValuesFrom ifc:LIST_1_3_IfcLengthMeasure ] ;
    # only for Extended layer
    rdfs:subClassOf [ a owl:Restriction; owl:onProperty ifc:coordinates;
                      owl:cardinality "1"^^xsd:integer ] .

@@

\subsubsection{Aggregated Types} may be unordered (\name{SET} and \name{BAG}) or  ordered (\name{LIST} and \name{ARRAY}) collections. \name{BAG} may contain duplicated values, unlike \name{SET}. \name{ARRAY} is specified with start and end indexes while other types are specified with min and max cardinalities.

\begin{principle}Any aggregated type must be subclass of one of \name{expr:SetClass}, \name{expr:ListClass}, \name{expr:ArrayClass} and \name{expr:BagClass} which are subclasses of \name{expr:\-Collec\-tion\-Class}. The latter has one a nonfunctional property \name{expr:slot} with type \name{expr:Slot} and four functional properties: \name{expr:isOrdered}, \name{expr:item\-Type}, \name{expr:\-size}, \name{expr:\-start\-Index}, \name{expr:\-end\-Index}. Class \name{expr:Slot} has two functional properties \name{expr:item} (the actual collection item value), \name{expr:index} and also optional functional properties \name{expr:previous}, \name{expr:next}. Below is a fragment of definition of type \name{SET [2:?] OF IfcProfileDef}:
\end{principle}

% In order to declare these types, new classes were created in a similar way as Ordered List Ontoloty \cite{olo:ontology}:
% \begin{itemize}
% \item Class \name{expr:\-Collec\-tion\-Class} has four functional properties: \name{expr:isOrdered}, \name{expr:item\-Type}, \name{expr:\-size}, \name{expr:\-start\-Index}, \name{expr:\-end\-Index} and a nonfunctional property \name{expr:slot} with type \name{expr:Slot}.
% \item Class \name{expr:Slot} has two functional properties \name{expr:item} (the actual collection item value), \name{expr:index} and also optional functional properties \name{expr:previous}, \name{expr:next}.
% \item Classes \name{expr:SetClass}, \name{expr:ListClass}, \name{expr:ArrayClass} and \name{expr:BagClass} are subclasses of \name{expr:\-Collec\-tion\-Class}.
% \end{itemize}

% Below is a fragment of definition of type \name{SET [2:?] OF IfcProfileDef}:

\begin{lstlisting}
# all layers:
 ifc:SET_2_UNDEFINED_OF_IfcProfileDef a owl:Class ;
    rdfs:subClassOf ifc:SET_2_UNDEFINED , ifc:SET_OF_IfcProfileDef .
 ifc:SET_2_UNDEFINED, SET_OF_IfcProfileDef rdfs:subClassOf expr:SetClass .

# only for layer Extended:
 ifc:SET_2_UNDEFINED rdfs:subClassOf rdfs:subClassOf [ a owl:Restriction ;
    owl:onProperty expr:slot ; owl:minCardinality "2"^^xsd:integer ] .
 ifc:SET_OF_IfcProfileDef rdfs:subClassOf [ a owl:Restriction ;
    owl:onProperty expr:slot ; owl:allValuesFrom ifc:SLOT_OF_IfcProfileDef ].
 ifc:SLOT_OF_IfcProfileDef rdfs:subClassOf [ a owl:Restriction ; owl:onProperty expr:item ; owl:allValuesFrom ifc:IfcProfileDef ] .
\end{lstlisting}

@@

\subsection{Generating ifcRDF Datasets From IFC Data}
\label{sec:ifcRDF}

\subsubsection{Dataset Metadata}
%\label{subsec:ifcRDF-metadata}

@@

\begin{lstlisting}
:GUID_1916794F-5605-4499-9AB0-F155DE8D3B6C
    a               ifc:IfcPropertySet ; # entity type
    ifc:globalId    [ a ifc:IfcGloballyUniqueId ; # defined type
                     rdf:value "0P5dbFLWL4cPgmyLNUZJji"^^xsd:string ] ;
    ifc:hasProperties  _:b11 .
_:b11 a ifc:SET_1_UNLIMITED_OF_IfcProperty ; # entity type
    expr:size "3"^^xsd:integer ;
    expr:itemType ifc:IfcProperty ;
    expr:slot _:b12, _:b13, _:b14 .
_:b12 a ifc:SLOT_OF_IfcProperty ;
    expr:item [ a ifc:IfcPropertySingleValue ; # entity type
                ifc:name "initial_GUID"^^xsd:string ;
                ifc:nominalValue  _:b15 ] .
\end{lstlisting}
